Local Type Inference for Polarised System F with Existentials

### 2024-04-30 oopsla24/tex/03-declarative-system.tex
Edit

The declarative system serves as a specification of the
type inference algorithm. It consists of two main parts:
the subtyping and the type inference.

\subsection{Subtyping}
It is represented by a set of inference rules shown in
\cref{fig:declarative-subtyping}.

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnDOneNsubLabeled{}

    \ottdefnDOnePsupLabeled{}
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
    \ottdefnDOneNeqLabeled{}

    \ottdefnDOnePeqLabeled{}
  \end{multicols}
  \caption{Declarative Subtyping}
  \label{fig:declarative-subtyping}
\end{figure}

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
the subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of the subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively.

@@

\paragraph{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that the subtyping on shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}.
Relaxing both of these invariants make the system equivalent to \systemf, and
thus, undecidable. However, after certain changes
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form, thereby
increasing the expressiveness of the system. These changes are discussed in
\cref{sec:weakening-invariant}.

\paragraph{Functions}
Standardly, the subtyping of function types is covariant in the return type
and contravariant in the argument type.

\paragraph{Variables}
The subtyping of variables is defined reflexively,
which is enough to ensure the reflexivity of subtyping in general.
The algorithm will use the fact that the subtypes of a variable 
coincide with its supertypes, which however is not true for an
arbitrary type.

\subsubsection{Properties of the Declarative Subtyping}
\label{sec:decl-subtyping-properties}

@@

\begin{property}[Subtyping Preserves Free Variables]
  \label{prop:subtyping-preserves-fv}
  Let us assume that all the mentioned types are well-formed in $[[Γ]]$. Then
  $[[Γ ⊢ iN1 ≤ iN2]]$ implies $[[fv(iN1) ⊆ fv(iN2)]]$,
  and $[[Γ ⊢ iP1 ≥ iP2]]$ implies $[[fv(iP1) ⊆ fv(iP2)]]$.
\end{property}

Another property that we extensively use is that the subtyping is reflexive and transitive,
and agrees with substitution.

\begin{property}[Subtyping forms a preorder]
  Let us say that two types $[[iN1]]$ and $[[iN2]]$ are in the subtyping relation if there 
  exists a context $[[Γ]]$ such that $[[Γ ⊢ iN1 ≤ iN2]]$;
  symmetrically, two types $[[iP1]]$ and $[[iP2]]$ are in the subtyping relation
  if there exists $[[Γ]]$ such that $[[G ⊢ iP2 ≥ iP1]]$.
  Then the subtyping relation defined this way is reflexive and transitive.
\end{property}

@@

Moreover, any two \emph{positive} types have the least upper bound, which makes
the positive subtyping semilattice.
The positive least upper bound can be found algorithmically,
which we will discuss in the next section.

\begin{property}[Positive Least Upper Bound exists]
  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
  well-formed in $[[Γ]]$.
  Then there exists the least common supertype--- a type $[[iP]]$ such that
  \begin{itemize}
    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
      $[[Γ ⊢ iQ ≥ iP]]$.
  \end{itemize}
\end{property}

\paragraph{Negative Greatest Lower Bound does not exist}
However, the symmetric construction---the greatest lower bound of two negative types
does not always exist. Let us consider the following counterexample.
Let us consider the following types: 
\begin{itemize}
  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
    none of the types is equivalent to $[[iQ]]$.
\end{itemize}
What is the greatest common subtype of  
$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
One of the common subtypes is $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$,
which, however is not the greatest one.

@@

By analyzing the inference rules, we can prove that
both $[[iM1]]$ and $[[iM2]]$ are maximal common subtypes.
Since $[[iM1]]$ and $[[iM2]]$ are not equivalent,
it means that none of them is the greatest.

@@

In our system, equivalence is reacher than equality. Specifically,
\begin{enumerate}
  \item[(ii)] one can introduce redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
  but not equal to $[[∀α⁺.↑α⁺]]$;
  \item[(i)] one can reorder quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
  $[[∀α⁺,β⁺.β⁺ → α⁺ → γ⁻]]$; 
  \item[(iii)] the transformations (i) and (ii) can happen at any
    position in the type.
\end{enumerate}

@@

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the \emph{ordering} function
$[[ord varset in iN]]$ and $[[ord varset in iP]]$.
For a set of variables $[[varset]]$, and a type, it 
returns a list of variables from $[[varset]]$ that occur in the type in 
the order of their first occurrence.
For brevity, we omit the formal definition of ordering,
referring the reader to the appendix.

@@

  \begin{multicols}{2}
  \ottdefnNrmNNormLabeled{}
  \\
  $[[ord varset in iN]]$ returns a list of variables
  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
  \columnbreak\\
  \ottdefnNrmPNormLabeled{}
  \\
  $[[ord varset in iP]]$ returns a list of variables
  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
  \end{multicols}

@@

For the normalization procedure, 
we prove soundness and completeness \wrt the equivalence relation.
\begin{property}[Correctness of normalization]
  \hfill \\
  \begin{itemize}
    \item[$-$] For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$;
    \item[$+$] analogously, for $[[iP]]$ and $[[iQ]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
  \end{itemize}
\end{property}

\subsection{Type Inference}

The declarative specification of the type inference is shown in 
\cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to have the type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.

The \emph{Application typing} judgment
infers the type of the application of a function to a list of arguments.
It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to have the type $[[iM]]$''.

@@

\paragraph{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows to infer
  the type of a variable from the context. 
  In literature can be found another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph{Annotations}
  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
  inferred type is a subtype of the annotation type.

@@

\paragraph{Return and Thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  add the corresponding shifts to the type of the body.

@@

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\paragraph{Typing up to Equivalence}
  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  They allow one to replace the inferred type with an equivalent one.

\paragraph{Application to an Empty List of Arguments}
  The base case of the application type inference is 
  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
  If the head of the type $[[iN]]$ is applied to no arguments, 
  the type of the result is allowed to be $[[iN]]$ or any 
  equivalent type. We need to relax this rule up to equivalence
  to ensure the corresponding property globally:
  the inferred application type can be replaced with an equivalent one.
  Alternatively, we could have added a separate rule similar to 
  \ruleref{\ottdruleDTPEquivLabel}, however, 
  the local relaxation is sufficient to prove the global property.

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  not known a priori. The algorithm we present in \cref{sec:algorithm} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

@@

\paragraph{Application of an Arrow Type}
  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

\subsubsection{Declarative Typing Properties}
  An important property that the declarative system has is
  that the declarative specification is correctly defined for
  equivalence classes.

\begin{property}[Declarative Typing is Defined up to Equivalence]
  Let us assume that $[[Γ ⊢ Φ1 ≈ Φ2]]$, \ie  
  the corresponding types assigned by $[[Φ1]]$ and
  $[[Φ2]]$ are equivalent in $[[Γ]]$.
  Also, let us assume that 
  $[[Γ ⊢ iN1 ≈ iN2]]$, $[[Γ ⊢ iP1 ≈ iP2]]$,
  and $[[Γ ⊢ iM1 ≈ iM2]]$. Then
  \begin{itemize}
    \item [$-$] $[[Γ ; Φ1 ⊢ c : iN1]]$ holds if and only if $[[Γ ; Φ2 ⊢ c : iN2]]$,
    \item [$+$] $[[Γ ; Φ1 ⊢ v : iP1]]$ holds if and only if $[[Γ ; Φ2 ⊢ v : iP2]]$, and
    \item [$\bullet$] $[[Γ; Φ1 ⊢ iN1 ● args ⇒> iM1]]$ holds if and only if $[[Γ; Φ2 ⊢ iN2 ● args ⇒> iM2]]$.
  \end{itemize}
\end{property}

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
The choice of polarization affects the execution strategy of the program.
We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
This choice makes it smoother to lift the translation to the term level, which will be disused next.

\begin{figure}[ht]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

@@

The term-level translations between systems are more complex.
They are defined as \emph{elaborations} of the typing derivations.
In other words, the input of the translation is not a term, 
but a whole typing derivation of this term in the corresponding system.
For brevity, we omit the formal definition of the term-level translation,
referring the reader to the appendix (\cref{sec:term-translation}).

We prove the soundness of the translation (\cref{lemma:translation-soundness}),
which in particular guarantees that the translation preserves typing.

\begin{property}[Soundness of the Translation]
    \label{lemma:translation-soundness}
    \begin{align*}
      & [[Γ ⊢ iN ≤ iM ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iN| → |iM| ]], \\
      & [[Γ ⊢ iP ≥ iQ ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iQ| → |iP| ]], \\
      & [[Γ; Φ ⊢ v : iP ⤳ ft]]  \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iP|]], \\
      & [[Γ; Φ ⊢ c : iN ⤳ ft]] \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iN|]], \\
      & [[fΓ ; fΦ ⊢ ft : fT ⤳ c]]  \text{ implies }  [[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]].
    \end{align*}
\end{property}

>>>>

In this section, we present the declarative system of \fexists,
which serves as a specification of the type inference algorithm.
The declarative system is represented as a set of inference rules
and consists of two main subsystems: subtyping and type inference.
First, we present the declarative subtyping rules specifying 
when one type is a subtype of another; next, we discuss 
the equivalence relation induced by mutual subtyping;
finally, we present the type inference rules,
that refer to the subtyping and equivalence relation.
We conclude this section by discussing the relation between the
proposed type system and the standard \systemf.

\subsection{Subtyping}

\begin{figure}
  \begin{minipage}[t]{0.49\textwidth}
    \ottdefnDOneNsubLabeled{}
  \end{minipage}%
  \begin{minipage}[t]{0.49\textwidth}
    \ottdefnDOnePsupLabeled{}
  \end{minipage}
  \hfill\\
  \begin{minipage}[t]{0.49\textwidth}
    \ottdefnDOneNeqLabeled{}
  \end{minipage}%
  \begin{minipage}[t]{0.49\textwidth}
    \ottdefnDOnePeqLabeled{}
  \end{minipage}
  \caption{Declarative Subtyping}
  \label{fig:declarative-subtyping}
\end{figure}

The inference rules representing declarative subtyping are shown in
\cref{fig:declarative-subtyping}. Let us discuss them in more detail.

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
This is the point of non-determinism as the instantiation can be chosen arbitrarily

@@

\paragraph{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, after
certain changes \ruleref{\ottdruleDOneShiftULabel} can be relaxed to the
covariant form, thereby increasing the expressiveness of the system. These
changes are discussed in \cref{sec:weakening-invariant}.

\paragraph{Functions}
Standardly, subtyping of function types is covariant in the return type
and contravariant in the argument type.

\paragraph{Variables}
Subtyping of variables is defined reflexively, which is enough to ensure the
reflexivity of subtyping in general. The algorithm---specifically the least
upper bound procedure---will use the fact that the subtypes of a variable
coincide with its supertypes (\cref{prop:var-no-subtypes}),
which however does not hold for arbitrary types.

\subsubsection{Properties of Declarative Subtyping}
\label{sec:decl-subtyping-properties}

@@

\begin{property}[Subtyping preserves free variables]
  \label{prop:subtyping-preserves-fv}
  Let us assume that all the mentioned types are well-formed in $[[Γ]]$. Then
  $[[Γ ⊢ iN1 ≤ iN2]]$ implies $[[fv(iN1) ⊆ fv(iN2)]]$,
  and $[[Γ ⊢ iP1 ≥ iP2]]$ implies $[[fv(iP1) ⊆ fv(iP2)]]$.
\end{property}

\begin{property}[Variable subtyping is trivial]
  \label{prop:var-no-subtypes}
  A subtype or a supertype of a variable is equivalent to the variable itself:

  \begin{tabular}{@{}llclcl@{}}
    $-$ & $[[Γ ⊢ α⁻ ≤ iN]]$ & $\iff$ & $[[Γ ⊢ iN ≤ α⁻]]$ & $\iff$ & $[[iN]] = [[∀pbs.α⁻]]$ \\
    $+$ & $[[Γ ⊢ α⁺ ≥ iP]]$ & $\iff$ & $[[Γ ⊢ iP ≥ α⁺]]$ & $\iff$ & $[[iP]] = [[∃nbs.α⁺]]$
  \end{tabular}
\end{property}

Another property that we extensively use is that subtyping is reflexive and transitive,
and agrees with substitution.

\begin{property}[Subtyping forms a preorder]
  For a fixed context $[[Γ]]$, the negative subtyping relation 
  $[[Γ ⊢ iN1 ≤ iN2]]$ and the positive subtyping relation 
  $[[Γ ⊢ iP1 ≥ iP2]]$
  are reflexive and transitive on types well-formed in $[[Γ]]$.
\end{property}

@@

Moreover, any two \emph{positive} types have a least upper bound, which makes
positive subtyping a semilattice. The positive least upper bound can be found
algorithmically, which we will discuss in the next section.

\begin{property}[Positive least upper bound exists]
  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
  well-formed in $[[Γ]]$.
  Then there exists a least common supertype--- a type $[[iP]]$ such that
  \begin{itemize}
    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
      $[[Γ ⊢ iQ ≥ iP]]$.
  \end{itemize}
\end{property}

\paragraph{Negative greatest lower bound might not exist}
The symmetric construction---the greatest lower bound of two negative types---does 
not always exist, as the following counterexample shows.
Consider the following types: 
\begin{itemize}
  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
    none of the types is equivalent to $[[iQ]]$.
\end{itemize}
What is the greatest common subtype of  
$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
The type $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$ is a common subtype,
however, it is not the greatest one, as it is too general.

@@

By analyzing the inference rules, we can prove that
both $[[iM1]]$ and $[[iM2]]$ are \emph{maximal} common 
subtypes, \ie there is no common subtype that is greater than them.
However, $[[iM1]]$ and $[[iM2]]$ are not equivalent,
which means that none of them is the greatest.

@@

In our system, equivalence is richer than equality. Specifically,
while staying within one equivalence class, one can 
change the type:
\begin{enumerate}
  \item[(i)] introduce and remove redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
  but not equal to $[[∀α⁺.↑α⁺]]$;
  \item[(ii)] reorder adjacent quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
  $[[∀α⁺,β⁺.β⁺ → α⁺ → γ⁻]]$; 
  \item[(iii)] make the transformations (i) and (ii) at any position in the type.
\end{enumerate}

@@

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the \emph{ordering} function
`$[[ord varset in iN]]$' and `$[[ord varset in iP]]$'.
For a set of variables $[[varset]]$, and a type, it 
returns a list of variables from $[[varset]]$ that occur in the type in 
the \emph{order of their first occurrence}.
For brevity, we omit the formal definition of ordering referring the reader to the appendix (\cref{sec:ordering}).

@@

  \begin{multicols}{2}
  \ottdefnNrmNNormLabeled{}
  \\
  $[[ord varset in iN]]$ returns a list of variables from 
  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
  \columnbreak\\
  \ottdefnNrmPNormLabeled{}
  \\
  $[[ord varset in iP]]$ returns a list of variables from 
  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
  \end{multicols}

@@

For the normalization procedure, 
we prove soundness and completeness \wrt the equivalence relation.
\begin{property}[Correctness of normalization]
  Assuming all types are well-formed in $[[Γ]]$,
  \begin{itemize}
    \item[$-$] $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$, and
    \item[$+$] $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
  \end{itemize}
\end{property}

\subsection{Typing}

The declarative specification of the type inference is given in 
\cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to infer type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.

The \emph{Application typing} judgment
infers the type of the application of a function to a list of arguments.
It has form $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to infer type $[[iM]]$''.

@@

\paragraph{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
  the type of a variable from the context. 
  In literature, one can find another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

@@

\paragraph{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

@@

\paragraph{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\paragraph{Typing up to equivalence}
  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  They allow one to replace the inferred type with an equivalent one.

\paragraph{Application to an empty list of arguments}
  The base case of the application type inference is 
  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
  If the head of the type $[[iN]]$ is applied to no arguments, 
  the type of the result is allowed to be $[[iN]]$ or any 
  equivalent type. We need to relax this rule up to equivalence
  to ensure the corresponding property globally:
  the inferred application type can be replaced with an equivalent one.
  Alternatively, we could have added a separate rule similar to 
  \ruleref{\ottdruleDTPEquivLabel}, however, 
  the local relaxation is sufficient to prove the global property.

\paragraph{Application of a polymorphic type $[[∀]]$}
  The complexity of the system lives in the rules whose output type is not
  immediately defined by their input and the output of their premises (\aka not
  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
  quantifying variables is not known a priori. The algorithm we present in
  \cref{sec:algorithm} delays this instantiation until more information about it
  (in particular, typing constraints) is collected.

@@

\paragraph{Application of an arrow type}
  Another application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

  \vspace{\baselineskip}
% \subsubsection{Properties of Declarative Typing}
  An important property that the declarative system has is
  that the declarative specification is correctly defined for
  equivalence classes.

\begin{property}[Declarative typing is defined up to equivalence]
  Let us assume that $[[Γ ⊢ Φ1 ≈ Φ2]]$, \ie  
  the corresponding types assigned by $[[Φ1]]$ and
  $[[Φ2]]$ are equivalent in $[[Γ]]$.
  Also, let us assume that 
  $[[Γ ⊢ iN1 ≈ iN2]]$, $[[Γ ⊢ iP1 ≈ iP2]]$,
  and $[[Γ ⊢ iM1 ≈ iM2]]$. Then
  \begin{itemize}
    \item [$-$] $[[Γ ; Φ1 ⊢ c : iN1]]$ holds if and only if $[[Γ ; Φ2 ⊢ c : iN2]]$,
    \item [$+$] $[[Γ ; Φ1 ⊢ v : iP1]]$ holds if and only if $[[Γ ; Φ2 ⊢ v : iP2]]$, and
    \item [$\bullet$] $[[Γ; Φ1 ⊢ iN1 ● args ⇒> iM1]]$ holds if and only if $[[Γ; Φ2 ⊢ iN2 ● args ⇒> iM2]]$.
  \end{itemize}
\end{property}

@@

The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
The choice of polarization affects the execution strategy of the program.
\ilyam{How?}
We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
This choice makes it smoother to lift the translation to the term level, which will be disused next.

The type-level translation is naturally lifted to the contexts:
$[[|Γ|]]$ is the context $[[Γ]]$ with every type depolarized,
and symmetrically, $[[+|fΓ|+]]$ is the context $[[fΓ]]$ with every type polarized.
The context translation is used to formulate the soundness of the
 term-level translation, which we discuss next.

\begin{figure}[ht]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization \protect\footnotemark}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}

\end{figure}

\footnotetext{Here the \systemf existential $[[∃ as . fT]]$ is a syntax sugar for its standard encoding $[[∀b.(∀as.(fT → b)) → b]]$}

@@

The term-level translations between the systems are more complex. They are
defined as \emph{elaborations} of the typing derivations. In other words, the
input of the translation is not a term, but a whole typing derivation in the
corresponding system. For example, the subtyping elaboration $[[Γ ⊢ iN ≤ iM ⤳
ft]]$ constructs a \systemf function $[[ft]]$ of type $[[|iN| → |iM|]]$ that
witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$; a typing elaboration
$[[Γ; Φ ⊢ v : iP ⤳ ft]]$ builds a term $[[ft]]$---a \systemf counterpart of
$[[v]]$---of type $[[|iP|]]$.

For brevity, we omit the formal definition of the term-level elaboration,
referring the reader to the appendix (\cref{sec:term-translation}). We define
the elaboration in such a way that it preserves the soundness invariants given
in \cref{lemma:translation-soundness}.

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
\end{table}

The least obvious aspect of the elaboration soundness is the application typing.
The elaboration judgment $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ 
outputs two things: 
\begin{enumerate*}
  \item [(i)] $[[fargs]]$---the \systemf counterpart of the argument list $[[args]]$, and 
  \item [(ii)] $[[fe]]$---a \systemf term that is used for the final cast 
    of the application $[[(x ● ft *)]]$ to the resulting type $[[|iM|]]$.
\end{enumerate*}
This final cast is needed to 
reflect the \fexists feature of \ruleref{\ottdruleDTEmptyAppLabel}
that permits the equivalent conversion of the resulting type $[[Γ ⊢ iN ≈ iN']]$,
as \fexists-equivalent types
might have different \systemf representations (\ie $[[|iN|]] \neq [[|iN'|]]$).
